One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-04 paper/discussion.tex
minor updates

\subsection{Unstable interaction between low-frequency $m=1$ spirals
  and the background disc due to an imposed temperature gradient}
The $m=1$ spiral that emerges in our simulations have low frequency
and grows slowly, and its spatial form is consistent local theory
(\S\ref{fargo_m1}). To demonstrate possible instability, we
first construct a neutral low-frequency mode using local theory, then
show that the torque density acting on this mode due to the background
temperature gradient is negative, which enforces the mode.

@@

%torque effective for large scales or sharp gradients

%Then, since $dc_s^2/dR < 0$, $T_\mathrm{BG}<0$ so the background
%disc applies a negative torque on the low-frequency $m=1$ spiral,
%which has negative angular momentum. This suggests the spiral
%amplitude will grow.

\subsection{Role of disc structure}

>>>>

\subsection{Unstable interaction between low-frequency $m=1$ spirals
  and the background disc due to an imposed temperature gradient}
The $m=1$ spiral that emerges in our simulations have low frequency
and grows slowly, and its spatial form is consistent local theory
(\S\ref{fargo_m1}). To demonstrate possible instability, we
first construct a neutral low-frequency mode using local theory, then
show that the torque density acting on this mode due to the background
temperature gradient is negative, which enforces the mode. % Note that
% we will not be assuming a self-gravitating disc until the end of this
% section.

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
the growth rate $\gamma$ of linear perturbations as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadradtic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
    + \ii\xi_\phi|^2}.
\end{align}
Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
in the local approximation, we have
\begin{align}
  2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
    \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
inequality. For $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ adopted in our disc
models,
\begin{align}
  2\gamma \lesssim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
\end{align}
Now we consider perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$ (as seen in
our simulations). Then
\begin{align}
  \gamma \lesssim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega = O(10^{-2})$,
consistent with simulation results.

Of course, we have been assuming a neutral mode to obtain expressions
for $\jlin$ and $T_\mathrm{BG}$ in the first place, so the above can only be
considered as an order-of-magnitude estimate. Nevertheless, the
expected growth rate is in reasonable agreement with numerical
results.

% \subsection{Role of self-gravity and disc structure} 
% The background torque density $T_{BG}$ does not depend on
% self-gravity. 
% %and by definition, disc structure plays no role in the local description of linear modes 
% However, our numerical simulations show that the $m=1$ spiral is
% confined between two $Q$-barriers, where real solutions to the local
% dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
% structure. Thus, the role of disc structure and self-gravity is to
% allow a neutral mode to be set up and confined, which is then
% destabilised through the background torque density.

%talk about additional simulations at high Q
%spiral mode as external potential to wider disc (lp11)

### 2014-11-04 paper/model.tex
minor updates

\section{Governing equations and numerical models}\label{model} 
The system of interest is a self-gravitating, inviscid fluid disc
orbting a central star of mass $M_*$. We will mainly consider two-dimensional
(2D, or razor-thin) discs in favour of numerical
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Hence, for generality we describe the system in
3D. We use both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally 
isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by 
\begin{align}\label{sound-speed}
  c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
\end{align}
where $h$ is the disc aspect-ratio $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient. A strictly isothermal disc
corresponds to $q=0$, and a disc with constant aspect-ratio
corresponds to $q=1$.

>>>>

\section{Governing equations and numerical models}\label{model} 
The system of interest is a self-gravitating, inviscid fluid disc
orbting a central star of mass $M_*$. We will mainly consider two-dimensional
(2D, or razor-thin) discs in favour of numerical
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Hence, for generality we describe the system in
3D. We use both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally 
isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by 
\begin{align}\label{sound-speed}
%  c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio and $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient. A strictly isothermal disc
corresponds to $q=0$, and a disc with constant aspect-ratio
corresponds to $q=1$.

### 2014-11-04 paper/results_3d.tex
minor updates

\section{Three-dimensional simulations}
In this section we briefly review results from 3D simulations carried
out using the ZEUS-MP and PLUTO codes. The main purpose is to confirm
the physical reality of the

@@

Snapshots at the end of the simulations are shown in
Fig. \ref{3d_prelim} for both ZEUS-MP and PLUTO. We again find the dead zone eventually
dominated by an $m=1$ spiral. A Cartesian visualisation of the
midplane density perturbation associated with the $m=1$ mode is shown in Fig. \ref{pluto_cart}.
The two codes give qualitatively similar results, but a longer PLUTO simulation was
needed to achieve similar mode amplitudes than ZEUS-MP. 
The amplitude of the spiral is smaller than the 2D case above,
possibily due to lower resolution and/or 3D self-gravity being weaker
than that in 2D.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/pdiskxy_020_m1}
  \caption{Cartesian visualisation the $m=1$ midplane density in the PLUTO
    simulation shown in Fig. \ref{3d_prelim}.\label{pluto_cart}}   
\end{figure}

>>>>

\section{Three-dimensional simulations}
In this section we review results from 3D simulations carried
out using the ZEUS-MP and PLUTO codes.

@@

Snapshots at the end of the simulations are shown in
Fig. \ref{3d_prelim} for both ZEUS-MP and PLUTO. We again find the dead zone eventually
dominated by an $m=1$ spiral.%  A Cartesian visualisation of the
% midplane density perturbation associated with the $m=1$ mode is shown in Fig. \ref{pluto_cart}.
The two codes give qualitatively similar results, but a longer PLUTO simulation was
needed to achieve similar mode amplitudes than ZEUS-MP. 
The amplitude of the spiral is smaller than the 2D case above,
possibily due to lower resolution and/or 3D self-gravity being weaker
than that in 2D.

@@

% \begin{figure}
%   \includegraphics[width=\linewidth]{figures/pdiskxy_020_m1}
%   \caption{Cartesian visualisation the $m=1$ midplane density in the PLUTO
%     simulation shown in Fig. \ref{3d_prelim}.\label{pluto_cart}}   
% \end{figure}
